Calculating life expectancy and exploring missingness

### 2016-09-15 Code/1_Explore_Data_v2.Rmd
first commit

---
title: "Exploratory Data Analysis -- Supressed Code of Death"
author: "Corinne Riddell"
date: "July 22, 2016"
output: html_document
---

@@

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?
```

@@

```{r calc_pop0}
View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1")])
timer2 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[2789]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

@@

This takes `r time1[3]` seconds to run.

@@

@@

Notes on Figure 1 (below):

@@

bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)

ggplot(bystate, aes(x = Age2, y = prop.missing)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")
```

@@

```{r save_workspace}
save.image(file = "/Users/corinneriddell/Dropbox/SRJ/Data/corinne/Sep5_workspace_bw_gap.Rdata")
```

>>>>

---
title: "Calculating life expectancy and exploring missingness"
author: "Corinne Riddell"
date: "July 22, 2016"
output: html_document
---

@@

summary(dat.clean$Crude_Rate[dat.clean$Population == 0])
table(dat.clean$Crude_Rate[dat.clean$Population==0], useNA = "always")
#whenever the population is 0, the Crude_Rate is equal to NA.
#need to think: should these be equal to 0, or NA? How will it affect the calculations?

#States with at least one strata of 0
na.pops <- dat.clean %>% select(State2, Population) %>% filter(Population == 0)
unique(na.pops$State2)

pops.le.20 <- dat.clean %>% select(State2, Population) %>% filter(Population <= 20)
unique(pops.le.20$State2)
```

@@

```{r calc_pop0, eval=F, echo=F}
View(dat2[dat2$StateYearRaceSex == levels(dat2$StateYearRaceSex)[2789], c("StateYearRaceSex", "Age2", "Population", "total_deaths1", 
                                                                       "Death_Rate1", "Ave_Lived_by_Died", "numYrsInt", "Prob_Dying1", "Prob_Surv1")])
timer2 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[2789]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

@@

```{r first_thousand}
timer3 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[6:1000]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

timer3
```

@@

```{r save_workspace1}
save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
```

```{r rest}
timer4 <- system.time (
  for(x in levels(dat2$StateYearRaceSex)[1001:9180]) {
      for(row in 2:length(dat2$alive_at_start_int1[dat2$StateYearRaceSex == x])) { 
        dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int1[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int1[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int5[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int5[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_int9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_int9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intb9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intb9[dat2$StateYearRaceSex == x][row - 1]
        dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row] <- dat2$alive_at_start_intw9[dat2$StateYearRaceSex == x][row - 1] - dat2$deaths_during_intw9[dat2$StateYearRaceSex == x][row - 1]

timer4
```

```{r rest_of_life_expectancy_calc2}
dat2$num_PY_int1 <- (dat2$alive_at_start_int1 - dat2$deaths_during_int1)*dat2$numYrsInt + (dat2$deaths_during_int1*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_int5 <- (dat2$alive_at_start_int5 - dat2$deaths_during_int5)*dat2$numYrsInt + (dat2$deaths_during_int5*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_int9 <- (dat2$alive_at_start_int9 - dat2$deaths_during_int9)*dat2$numYrsInt + (dat2$deaths_during_int9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_intb9 <- (dat2$alive_at_start_intb9 - dat2$deaths_during_intb9)*dat2$numYrsInt + (dat2$deaths_during_intb9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)
dat2$num_PY_intw9 <- (dat2$alive_at_start_intw9 - dat2$deaths_during_intw9)*dat2$numYrsInt + (dat2$deaths_during_intw9*dat2$Ave_Lived_by_Died*dat2$numYrsInt)

```{r save_workspace2}
save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
```

@@

###Initial graphs on missing data
Notes on Figure 1 (below):

@@

#calculate the size of the black population in 1969 by State
bystate.bpop.1969 <- 
  dat %>%
  filter(COD2 == "Cancers" & Year2 == 1969 & Race2 == "Black") %>% #to stop repeats of the population size
  group_by(State2) %>%
  summarize(pop_by_race_1969 = sum(as.numeric(Population)))

bystate <- merge(bystate, bystate.bpop, by = "State2", all.x = T)
bystate <- merge(bystate, bystate.bpop.1969, by = "State2", all.x = T)
bystate$State3 <- reorder(bystate$State2, bystate$pop_by_race)

ggplot(bystate, aes(x = Age2, y = prop.missing)) + 
  geom_point(aes(col=Race2)) + facet_wrap(~ State3, ncol = 4) +
  theme_bw(base_size = 11) +
  theme(legend.position="top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Proportion Missing (%)") +
  xlab("Age group")

ggplot(subset(bystate, Age2 == "<1 year"), aes(pop_by_race_1969, pop_by_race)) + geom_point() + geom_abline(intercept = 0, slope = 1)

unique(bystate$State2[bystate$Age2 == "<1 year" & bystate$pop_by_race_1969 < 10000])
```

@@

```{r save_workspace3}
save.image(file = "/Users/corinneriddell/Dropbox/BlackWhiteGap/Data/Sep6_BWgap.Rdata")
```
